\documentclass[12pt]{article}

% House style preamble
\input{.house-style/preamble.tex}

% Update PDF metadata
\hypersetup{
  pdftitle={Function-first without function-in-syntax: projectibility, field-relative kinds, and the limits of Thomasson's deflation},
  pdfkeywords={conceptual engineering, natural kinds, projectibility, deflationism, easy ontology, modal normativism}
}

\setcounter{biburlnumpenalty}{10000}
\setcounter{biburllcpenalty}{10000}
\setcounter{biburlucpenalty}{10000}
\AtBeginBibliography{\raggedright}

\title{Function-first without function-in-syntax\\
{\large Projectibility, field-relative kinds, and the limits of Thomasson's deflation}}
\author{Brett Reynolds \orcidlink{0000-0003-0073-7195}%
\thanks{Contact: \href{mailto:brett.reynolds@humber.ca}{brett.reynolds@humber.ca}}\\
Humber Polytechnic \& University of Toronto}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
Amie Thomasson's metaphysics treats ontological and modal disputes through rules of use, discourse roles, and conceptual revision. I accept this function-first turn and detach it from the systemic functional linguistics used to motivate it in \emph{Rethinking Metaphysics}. The needed premise is role pluralism: categories do different jobs in different practices.

Natural-kind cases show why role pluralism needs a projectibility constraint. The cases of \mention{fish}, \mention{jade}, and \mention{race} don't separate conceptual engineering from projectibility realism, since Thomasson can accommodate their empirical corrections. The harder case arises when a category keeps coordinating a field after losing the projection that gave it kind-standing. A projectibility realist account adds the missing demotion rule: once a field and projection are fixed, stable projection failure can defeat a useful category profile.
\end{abstract}

\noindent\textbf{Keywords:} conceptual engineering; natural kinds; projectibility; deflationism; easy ontology; modal normativism

\section{Introduction}

Thomasson's work gives metaphysics a functional turn. Existence questions are handled through application conditions for terms \citep{thomasson2014ontology}. Metaphysical modal claims express object-language rules for what can be said, inferred, and challenged in modal discourse, rather than descriptions of modal features of reality \citep{thomasson2020norms}. Conceptual engineering becomes central once we ask which concepts and categories should be retained, revised, or introduced \citep{thomasson2025rethink}. These claims don't eliminate metaphysics. They redirect it.

Projectibility doesn't undo that redirection. It disciplines it. By \term{deflation}, I mean not the elimination of metaphysics, but the denial that metaphysical questions are answered by a field-neutral inventory of privileged structures.

Here \term{function} means role in a practice, not causal mechanism, biological purpose, or syntactic slot. Four levels need to be kept distinct. At the semantic and conceptual level, terms have rules for correct application and inference. At the discourse level, an expression or claim can mark a commitment, license a contrast, repair disagreement, or organize inquiry.

At the field-inferential level, categories license projections and answer questions. At the testing level, those projections can be preserved, weakened, or defeated. The paper uses \term{function} across use, discourse, and inquiry, not as a claim about linguistic architecture.

Sections 2 and 3 use the semantic/conceptual and discourse levels to detach Thomasson's functional pluralism from systemic functional linguistics. Sections 4--6 use the field-inferential and testing levels to formulate the projectibility realist alternative.

A small linguistic example, anticipating the Hallidayan discussion, keeps the levels apart. \mention{London} can be treated as a \term{proper name} when the question concerns reference, opacity, or identity, and as a \term{proper noun} when the question concerns distribution, determiner resistance, agreement, or modification. The same string can have a referential role and a syntactic profile, but one isn't read off from the other.

A field-relative view also changes how ordinary discourse enters the argument. By a \term{field}, I mean a stabilized practice or inquiry with its own questions, norms, and defeaters. Ordinary linguistic practice can be treated as one such field. Metaphysics is another, including hard ontology understood as technical inquiry into what exists and how it's structured. Thomasson's arguments often let ordinary application conditions discipline metaphysical claims, especially when metaphysicians purport to use ordinary terms. That discipline is legitimate, but it doesn't decide in advance how technical or revisionary metaphysical vocabulary should be assessed.

Field individuation is epistemic rather than ontological. A field is stable enough for this argument when practitioners share questions, admissible evidence, defeaters, and repair moves. Its boundaries can shift, but shifting the field changes the projection under test; it doesn't protect a category from failure.

This paper accepts Thomasson's redirection. Its first claim is that her function-first metaphysics needs less linguistic machinery than she sometimes suggests. The philosophical result depends on a general point: different parts of discourse have different semantic, inferential, and sorting roles, and metaphysics often goes wrong by treating all of them as attempts to describe the world's basic inventory.

Natural kinds supply the second claim. Thomasson is at her strongest when addressing Kripke-Putnam essentialism. She has the resources to resist the thought that metaphysical inquiry discovers a privileged structure that any adequate category scheme has to mirror. But a large anti-essentialist tradition treats kinds through projectibility, causal support, plural categorization, and defeasible inductive use. That tradition is the realist alternative not yet central in her argument.

I defend a field-relative account of kind claims. A category has standing within a field when membership in it supports reliable projections: observing some features licenses expectations about others in ways that survive appropriate tests \citep{Goodman1955}. A kind claim then says that this standing is strong enough for kindhood in that field. This account grants Thomasson's anti-essentialism and functional pluralism. It denies that those concessions settle the realist question. The point isn't to choose between realism and pluralism. It's to distinguish conceptual-engineering pluralism from projectibility-constrained pluralism.

\section{Thomasson's function-first metaphysics}

In \emph{Ontology Made Easy}, Thomasson explains existence questions through application conditions. These are rules for competent use in discourse rather than definitions speakers have to state. Nor are they merely evidence conditions. They specify when applying a term would be correct, not simply when we'd be warranted in thinking the term applies \citep[93--95]{thomasson2014ontology}.

This machinery already accommodates more empirical deference than a simple conventionalist reading\ would suggest. Thomasson's \mention{platypus}\ example allows natural-kind terms to defer to empirical facts and to experts when biological distinctions are at issue \citep[94]{thomasson2014ontology}.

That doesn't mean a technical category automatically sets the standards for everyday discourse in every setting. It means that some ordinary natural-kind uses already include a norm of deference when biological correctness is at issue. So the objection can't be that Thomasson has no room for natural-kind terms, causal discovery, or expert correction.

\emph{Norms and Necessity} extends the same functional strategy to modal discourse. For metaphysical modal discourse, modal claims express rules of use rather than tracking independent modal features. Thomasson also separates that claim from causal, nomological, dispositional, and empirical counterfactual claims. At the field-inferential level, scientific laws can guide empirical inference, and natural-kind categories can be assessed by how well they support predictive and explanatory generalizations \citep[121--123]{thomasson2020norms}.

\emph{Rethinking Metaphysics} makes the positive program explicit. Conceptual engineering isn't arbitrary when it responds to empirical, technological, and social constraints. Thomasson's discussion of \mention{fish} is especially important. Biological concepts can change as empirical discoveries and theoretical purposes change; the resulting categories can be assessed by how well they support generalization and prediction \citep[216--217]{thomasson2025rethink}.

Her discussion of\ \mention{race} makes the same point from the other side. Identifying the functions of racial terms can reveal failed theoretical uses and objectionable social functions. So Thomasson already has room for failed projection to motivate conceptual change \citep[210--213]{thomasson2025rethink}.

Her\ \mention{jade} discussion makes the convergence still clearer. Following Rouse, Thomasson treats \mention{gold} and \mention{jade} as cases where empirical surprises force decisions about use that depend on what a practice is doing. In her own verdict, we reject iron pyrites as gold but accept two kinds of jade \citep[101--102]{thomasson2020norms}.

These passages create the central convergence. Thomasson already has an empirically constrained account of concept and category choice, responsive to the roles concepts play in use, discourse, inquiry, and revision. The question is whether that account exhausts what a realist projectibility theory wants from kinds.

The projectibility-first account should keep that lesson without making the route one-way. Kind inquiry can start from categories already in use or from a discovered profile of co-varying properties. What it can't start from is a field-neutral inventory of essences. In either direction, the kind claim earns standing only when the category and profile support projections within a field.

Projectibility isn't just another purpose a concept might serve. It's a way of testing whether a category's success outruns our endorsement of it. A projectible category licenses expectations that can be preserved, weakened, or defeated when conditions change. That gives kind claims a world-answerable dimension even when they remain plural and field-relative.

\section{The dispensable Hallidayan dependency}

\emph{Rethinking Metaphysics} gives systemic functional linguistics (SFL) a constructive role. Thomasson uses it to identify what linguistic forms do, and especially to explain why noun-term recastings are useful rather than ontologically suspect \citep[133--160]{thomasson2025rethink}.

Her central cases are simple. From \mention{The barn is red}, speakers can move to \mention{the barn has} [\mention{the property of redness}]; from\ \mention{there are two cups on the table}, they can move to [\mention{the number of cups on the table}]\ \mention{is two}. SFL helps explain why these shifts aren't idle ontological inflation: the noun-term versions let speakers refer back, qualify, quantify, generalize, and build theory with the recast material \citep[154--156, 169--179]{thomasson2025rethink}.

Systemic functional linguistics is a Hallidayan grammatical framework that treats linguistic form in relation to what language lets speakers do in social use. In that framework, discourse roles aren't a later pragmatic add-on to grammar. They belong inside the architecture of grammatical description.

That use of SFL motivates the dependency question. If Thomasson's metaphysical result depends on the Hallidayan placement of discourse roles inside grammar, then a general point about discourse roles has been tied to one grammatical architecture. An early review presses the same dependency worry from a hostile direction: SFL adds little to the explanation of how sentence content is encoded, leaving that work to compositional semantics \citep{williamson2026rethinking}.

\textcite{reynolds2026thomassonSFLReview} develops the linguistic version of the extraction argument: SFL supplies a useful diagnostic vocabulary, but the relevant discourse-role insight can be stated without adopting Hallidayan grammatical architecture. This paper asks what follows for kinds once that detachment is granted.

Only a modest premise is needed: expressions and claims can have different discourse jobs. Easy ontology and modal normativism already use role-sensitive resources, and inferentialist, pragmatist, or conceptual engineering vocabularies can state the same point. The barn and cup examples need the easy-ontology machinery plus an account of why the recast formulations are worth having; they don't need the stronger grammar.

For this paper, the detachment claim is limited. I am not arguing that the linguistic framing is useless. It may help Thomasson organize the view and connect it to a broader account of discourse. The claim is narrower: the metaphysical payoff comes from function-first pluralism, not from a grammar-internal account of function.

\section{A projectibility realist alternative}

Thomasson's main natural-kind focus is the essentialist tradition associated with Kripke and Putnam \citep{Kripke1980,Putnam1975Meaning}. That focus is important, but it isn't the whole debate. The anti-essentialist projectibility tradition isn't treated as the central realist alternative.

The omission matters because the anti-essentialist realist tradition grants much of what Thomasson wants. Goodman made projectibility central to induction: not every true predicate supports reliable projection \citep{Goodman1955}. His point can be put without the technical details of \mention{grue}: two predicates can fit all observed cases while only one supports new expectations. The difference lies in projectibility, not in past truth alone.

Dupré rejects the idea that objects generally have one context-independent natural kind and defends a pluralist realism on which overlapping categories can be equally real \citep[5--7, 57--58, 262]{dupre1993disorder}. Boyd's homeostatic property cluster account, Khalidi's causal-network account, Magnus's pluralist critique of reducing natural-kind realism to homeostatic property clusters, and Slater's \term{natural kindness} all weaken or reject essentialism while retaining a realist role for kinds \citep{boyd1991,boyd1999,khalidi2013,khalidi2018nodes,magnus2012scientific,slater2015}.

Boyd gives the clearest connection to Thomasson. On his accommodationist view, the reality of a kind turns on whether reference to it contributes, within a disciplinary setting, to reliable inference and accommodation to causal structure. Different species concepts can answer to different explanatory projects without becoming merely unreal or arbitrary \citep[169--176]{boyd1999}. Khalidi makes the parallel point in causal-network terms. Categories needn't be mind-independent; they need to serve epistemic purposes by tracking causal patterns and supporting projectible inferences \citep[202--203]{khalidi2013}.

Species concepts make the point concrete. In an ecosystem-oriented inquiry, a field biologist may care about organisms that occupy the same niche, compete for the same resources, host the same parasites, or play the same trophic role. That category should project interaction effects: if one population is removed, which relations in the system change? In a genetic inquiry, the central projections concern gene flow, reproductive isolation, ancestry, and hybridization. Two populations can be grouped together for ecological purposes and distinguished for genetic purposes, or vice versa, without either result being arbitrary.

Dupré still matters because he states the pluralist lesson bluntly. Category choice depends on the goals of inquiry, but the resulting kinds needn't be unreal. Species may be treated as individuals in some contexts and as modest kinds in others \citep[235]{dupre1993disorder}. He uses Goodmanian projectibility only in passing \citep[279 n.~20]{dupre1993disorder}, so he doesn't provide the full apparatus of graded support, demotion conditions, and field discipline that a projectibility-first account needs.

\section{Projectibility-first, field-relative kinds}

A projectibility-first account begins with the projection to be licensed. A category earns standing when membership in it supports reliable expectations about other features, behaviours, failure modes, or intervention outcomes. The first question isn't what essence all members share, but what inferences the category licenses, under what conditions, and with what known breakdown points.

This requires three distinctions. A category sorts. A profile projects. A kind claim says that the sorting earns standing because it tracks the profile well enough for a field. Confusing these three levels makes weak categories look like well-supported kinds.

Field-relativity here isn't whim-relativity. Fields have measurement routines, authorities, records, and defeaters. Field-relative kind claims can be answerable to the world because their projections can fail. Categories can also be demoted: a category may remain useful for one task while losing standing for another.

The apparatus has three parts. Graded support concerns the reliability, scope, and counterfactual robustness of the projections a category licenses. Failure conditions specify breakdown rather than noise: anomalous cases, field changes, intervention effects, or loss of explanatory integration.

For demotion, hold the field fixed. A category loses kind-standing for a projection when its sorting or coordination use can remain intact while the projection fails in stable, explainable ways, especially where a rival profile identifies the failures. The category may still serve communication, intervention, or record-keeping. What it loses is standing as a kind for that field's inferential task.

The \mention{fish} case shows the distinction. In ordinary food and market practices, \mention{fish} can be a useful category because it supports expectations about handling, preparation, and labelling. In phylogenetic systematics, the same category can be demoted because it doesn't track the ancestry relations that field asks about. In ecological work, a narrower category built around aquatic life may still support expectations about habitat, trophic role, or exposure. The word alone doesn't settle the kind claim; the field and projection do. A category can be good for the kitchen and bad for the cladogram.

That case doesn't yet separate the views. A useful label survives in one field while a projection fails in another, and Thomasson's framework can accommodate that. The same is true of \mention{jade}, since Thomasson accepts the empirically driven move to two kinds of jade \citep[101--102]{thomasson2020norms}. A separating case needs stricter field control: the issue can't be that a useful label survives in one field while a projection fails in another.

The sharper case holds the field fixed. Within one practice, the same category serves both a coordination role and an inferential role; the coordination role remains useful while the projection fails in stable, explainable ways. At that point, function-first assessment can point to continued usefulness, but it still needs a rule for which in-field role fixes kind-standing.

English nominal grammar uses \term{definiteness} both to coordinate diagnostics and to project a form--meaning package.

It also uses \term{definiteness} to teach article choice, organize noun-phrase diagnostics, connect descriptive grammars, compare semantic theories, and mark a shared problem-space for syntax and meaning. The inherited category carries a bundled projection. Formal signs of definiteness are expected to travel with identifiability, uniqueness, familiarity, anaphoric or deictic anchoring, existential-\mention{there} restrictions, partitive-\mention{of} eligibility, and related diagnostics \citep{lyons1999,reynolds2025definiteness}.

Stable dissociations show the projection failure. Weak definites such as \mention{take the bus} and \mention{go to the hospital} have formal definiteness without ordinary uniqueness \citep{carlsonsussman2005,aguilarguevarazwarts2010}. Generic definites such as \mention{the lion} support kind reference \citep{KrifkaEtAl1995}. Proper names can be semantically definite without the ordinary article profile. Bare institutional nominals such as \mention{in hospital} and \mention{go to school} can license institutional or definite-like readings without ordinary definite marking \citep{reynolds2025definiteness}. These are stable dissociations inside one field.

The definiteness case differs from that control. The unified label keeps coordination value after the projection failures are known. Researchers still use \term{definiteness} to connect diagnostics, exceptions, article-choice patterns, semantic theories, and pedagogical explanations. A function-first account therefore has reason to retain the unified category: it still organizes the field.

On the projectibility-first account, the unified profile is demoted for the form--meaning projection. The projection-bearing profiles are more specific: formal definiteness, anchored in morphosyntactic and distributional diagnostics, and semantic definiteness, anchored in identifiability, uniqueness, familiarity, and related pragmatic conditions. The category remains useful; the projection has split.

Because Boyd is the closest realist ally, projectibility-first realism might look like a relaxed version of his homeostatic property cluster view. That would be too narrow, but it also shouldn't become ground-free. Lipski's criticism of neo-essentialist readings of homeostatic property cluster theory and Slater's \term{natural kindness} both resist the thought that kindhood requires a uniform homeostatic mechanism \citep{lipski2020,slater2015}.

Martinez presses the other side of the dialectic. Against Slater's rejection of the grounding claim, he argues that grounds can be genuinely explanatory of the epistemic practices that make stable clusters scientifically useful \citep{martinez_2017_stable_property_clusters_grounds}. Projectibility-first realism should accept that constraint in field-indexed form. Grounds matter when they explain stable projection failures and identify why a rival profile does better.

What this paper adds is field-indexed demotion. A stable profile, a network position, a sustaining mechanism, and corrective control can come apart. Some projectible profiles are control-secured. Others draw support from history, isolation, background network structure, or institutional uptake. The question is which support source explains the projection and its failures.

\section{The residual disagreement}

The disagreement with Thomasson should be stated narrowly. Purposes matter; empirical facts constrain concept and category choice; revisionary metaphysics can have a legitimate role. Thomasson's \mention{jade} discussion shows that empirical surprises can force decisions about use \citep[101--102]{thomasson2020norms}, and her discussions of \mention{race} and \mention{fish} show that conceptual engineering can be worldly and non-arbitrary \citep[210--217]{thomasson2025rethink}. Her modal normativism also allows revisionary metaphysical views to be treated as proposals for new rules for key terms, assessed by the benefits of those changes \citep[205--206]{thomasson2020norms}.

A priority worry remains, but it should be put as a dilemma. If a metaphysician uses \mention{table} to make an ordinary existence claim, ordinary application conditions constrain the claim. If she introduces \mention{mereological fusion} or \mention{simples arranged as a table}, ordinary discourse no longer settles the issue. The technical vocabulary then has to earn its keep by explaining composition, persistence, counting, or modal contrasts that ordinary furniture talk doesn't supply.

The worry concerns transfer across fields. Once these practices are treated as fields of discourse and inquiry, usefulness in one leaves standing in another unsettled. Ordinary use has evidential force without final authority. Hard ontology gets a field-relative burden. Distance from ordinary discourse becomes a defect when the technical vocabulary fails to license projections, explanations, or disciplined contrasts.

Projectibility may still add a realist constraint beyond empirically informed usefulness. The remaining issue concerns demotion. A category can serve several functions at once: coordination, repair, pedagogy, record-keeping, and inference. When those roles pull apart, function-first assessment doesn't by itself say which role fixes kind-standing. Projectibility-first realism re-indexes the assessment to a specified projection. Failed projection isn't then one cost among all practical benefits. It defeats the kind claim for that field.

The original-purpose reply doesn't settle the matter. A term can enter a practice to mark an inference and later become the device by which the field coordinates exceptions, diagnostics, and repairs. At that point, the coordination role has its own standing in the practice. Function-first pluralism needs a further rule for choosing between the original inferential role and the later coordinating role. Projectibility-first realism chooses the projection explicitly.

Successful conceptual functioning is too coarse-grained to replace projectibility. A concept can coordinate inquiry, preserve continuity with a literature, support teaching, and organize records while failing the inference that would make it a kind for a specified projection. Projectibility-first realism grants those other successes. It denies that they compensate for projection failure when the kind claim is indexed to that projection.

Thomasson can respond that this still doesn't identify a field-neutral structure that mandates one correct category scheme. That response is right. The projectibility-first account needs only the weaker realist claim that, once a field and projection are specified, the world can make some category profiles better supported than others.

\section{Conclusion}

Thomasson's function-first metaphysics should be accepted in its strongest form. It redirects metaphysics toward rules of use, discourse roles, field-relative category choice, and conceptual revision. It also shouldn't be dismissed on the grounds that it can't handle natural-kind deference or empirical constraint. But ordinary linguistic practice by itself isn't a field-neutral court over every technical practice, including metaphysics itself.

The anti-essentialist realist tradition remains largely outside her focus. Projectibility-first realism accepts pluralism, rejects essences, and treats kind claims as field-relative. It then asks for graded support, failure conditions, and demotion rules. Thomasson's account hasn't yet shown that this demotion discipline can be absorbed into conceptual engineering alone.

Projectibility-first realism doesn't ask for a privileged category scheme fixed independently of inquiry. It asks what follows once a field, a category, and a projection are specified. Some categories then earn more than usefulness: they support defeasible expectations, explain failures, and generate tests that can be preserved or defeated by the world. That's the realist residue: not an essence behind the category, but a risk the category has to run.

\section*{Acknowledgements}

OpenAI Codex (GPT-5) served as a drafting and editing aid in the preparation of this paper. Anthropic Claude Opus 4.8 provided review and revision ideas. I am responsible for all theoretical claims, arguments, errors, and interpretive choices.

\clearpage
\printbibliography

\end{document}
